---
output:
  html_document: default
  pdf_document: default
---

Financial Contributions to Presidential Campaigns by State 
================================================================================

### By Ruchita Maheshwary

In this report, I analyzed the data set for around 1304500 contributions for 
presidential elections from the state of California.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gender)
library(tidyr)
library(data.table)
library(dplyr)
theme_update(axis.title=element_text(size=12),axis.text=element_text(size=10),
             plot.title=element_text(size=14))

```

```{r echo=FALSE, Load_the_Data}
# Load the Data with replacing blank values with N/A
df <-read.csv('P00000001-CA.csv',na.strings=c("", "N/A"),row.names = NULL)
#matching correct column names with the data
colnames(df) <- c(colnames(df)[-1],"x")
df$x <- NULL


```

**Brief description about the dataset variables:**

* COMMITTEE ID- A 9-character alpha-numeric code assigned to a committee by the 
Federal Election Commission. 

* CANDIDATE ID - A 9-character alpha-numeric code assigned to a candidate by the
Federal Election Commission. 

* CANDIDATE NAME

* CONTRIBUTOR NAME - Reported name of the contributor.

* CONTRIBUTOR CITY - Reported city of the contributor.

* CONTRIBUTOR STATE - Reported state of the contributor.

* CONTRIBUTOR ZIP CODE - Reported zip code of the contributor.

* CONTRIBUTOR EMPLOYER - Reported employer of the contributor.

* CONTRIBUTOR OCCUPATION - Reported occupation of the contributor.

* CONTRIBUTION RECEIPT AMOUNT - Reported contribution amount.

* CONTRIBUTION RECEIPT DATE - Reported contribution receipt date.  
  The date format is DD-MMM-YYYY.

* RECEIPT DESCRIPTION - Additional information reported by the committee about 
  a specific contribution. 

* MEMO CODE - 'X' indicates the reporting committee has provided additional text
to describe a specific contribution.  See the MEMO TEXT.  

* MEMO TEXT - Additional information reported by the committee about a specific 
contribution.  

* FORM TYPE - Indicates what schedule and line number the reporting committee
reported a specific transaction.<br/>   

    * SA17A:  Form 3P Schedule A Line 17A<br/>
    * SA18:   Form 3P Schedule A Line 18<br/>
    * SB28A:  Form 3P Schedule B Line 28A<br/>  

* FILE NUMBER - A unique number assigned to a report and all its associated 
transactions.   

* TRANSACTION ID - A unique identifier permanently associated with each
itemization or transaction appearing in an FEC electronic file.

* ELECTION TYPE / PRIMARY-GENERAL INDICATOR - This code indicates the election 
for which the contribution was made. EYYYY (election plus election year)<br/>

    * P = Primary<br/>
    * G = General<br/>
    * O = Other<br/>
    * C = Convention<br/>
    * R = Runoff<br/>
    * S = Special<br/>
    * E = Recount<br/>


# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}

dim(df)

str(df)

summary(df)

```

```{r echo=FALSE, load_first_last_name}
#Load candidates by first and last name
df$cand_nm.new <- as.character(df$cand_nm)
#split candidate name by first and last name and identify unique candidate names
#in the data set
ca_contrib <- extract(df, cand_nm.new, c("last_name","first_name"), 
                      "(.*),\\s*(\\S*)\\s*.*", fill="left",perl=TRUE)
ca_candidates <- unique(ca_contrib$first_name)

```

```{r echo=FALSE, candidate_gender}
#Load gender of candidates
#retrieve gender of the candidates with year of birth between 1941-1971
gender <- gender(as.character(ca_candidates), c(1941, 1971),method = "ssa",
                 countries = "United States")
#join with the data set the gender data and remove additional columns
ca_contrib<-left_join(x=ca_contrib,y=gender,by=c("first_name"= "name"))
names(ca_contrib)[names(ca_contrib) == 'gender'] <- 'candidate_gender'
drop <- c("name","proportion_male","proportion_female","year_min","year_max")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
ca_contrib$candidate_gender <- as.factor(ca_contrib$candidate_gender)
```

```{r echo=FALSE, load_contributor_details}
#Repeat the same process for contributors
ca_contrib$contrib_nm.new <- as.character(ca_contrib$contbr_nm)
#retrieve contributor first and last name
ca_contrib <- extract(ca_contrib, contrib_nm.new, c("contrib_last_name",
                                                    "contrib_first_name"),
                      "(.*),\\s*(\\S*)\\s*.*", fill="left",perl=TRUE)
ca_contributors <- unique(ca_contrib$contrib_first_name)
#retrieve gender of the contributors with year of birth between 1922-1997
gender_contrib <- gender(as.character(ca_contributors), c(1922, 1997),
                         method = "ssa",countries = "United States")
#join with the data set the gender data and remove additional columns
ca_contrib<-left_join(x=ca_contrib,y=gender_contrib,
                      by=c("contrib_first_name"= "name"))
names(ca_contrib)[names(ca_contrib) == 'gender'] <- 'contributor_gender'
drop <- c("contrib_last_name","contrib_first_name","name",
          "proportion_male","proportion_female","year_min","year_max")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
ca_contrib$contributor_gender <- as.factor(ca_contrib$contributor_gender)

```

```{r echo=FALSE, remove_columns}
#Removing non-usable columns from data frame like committe id,receipt 
#description and redundant columns like contributor state etc


#drop <- c("cmte_id","contbr_st","receipt_desc","memo_cd","memo_text","tran_id")
#ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]

```

```{r echo=FALSE, load_candidate_party}
#Add column of political party for each presidential election candidate based on
#candidate last name

candidateParty <- function(x) {
    if(is.na(x)){
      NA
    }else if(x == "Clinton" | x == "Sanders" | x == "O'Malley" | x=="Lessig" 
             | x=="Webb"){
      "Democratic"
    }else if(x == "Johnson"){
      "Libertarian"
    }else if(x=="Stein"){
      "Green"
    }else if(x=="McMullin") {
      "Independents"
    }else {
      "Republican"
    }
}

ca_contrib$party <- apply(ca_contrib['last_name'],1,candidateParty)

```

```{r echo=FALSE, load_election_type}

#Add election type description for "election_tp" column

electionType <- function(x) {
  if(is.na(x))  {
    NA
  }else if(substr(x, 1, 1)=="P")  {
    "Primary"
  }else if(substr(x, 1, 1)=="G")  {
    "General"
  }else if(substr(x, 1, 1)=="O")  {
    "Other"
  }else if(substr(x, 1, 1)=="C")  {
    "Convention"
  }else if(substr(x, 1, 1)=="R")  {
    "Runoff"
  }else if(substr(x, 1, 1)=="S")  {
    "Special"
  }else {
    "Recount"
  }
}

ca_contrib$election_type <- apply(ca_contrib['election_tp'],1,electionType)
drop <- c("election_tp")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
  
```


```{r echo=FALSE, retrieve_contribution_dates}

#Obtain contribution month/year
ca_contrib$contribution_date <- as.Date(ca_contrib$contb_receipt_dt,"%d-%B-%y")
ca_contrib$contrib_year = as.numeric(format(ca_contrib$contribution_date,
                                            format="%Y"))
ca_contrib$contrib_month = months(as.Date(ca_contrib$contribution_date))
#ca_contrib = ca_contrib[,!(names(ca_contrib) %in% c("contribution_date"))]

```


```{r  echo=FALSE, message=FALSE, warning=FALSE, contribution_amounts}
#On analyzing the data I found some negative contributions as well which were 
#basically refunds (typically following a candidate dropping out of the race). 
#So will subset the data into positive and negative contributions

#positive_ca_contrib_amt <-subset(ca_contrib,contb_receipt_amt>0)
#negative_ca_contrib_amt <-subset(ca_contrib,contb_receipt_amt<=0)

```

After cleaning the data our dataset contains 1304346 observations of 26 
variables.

I want to determine count of male/female candidates who contested the 
Presidential election. Plot for male/female candidate count in 
the presidential election
```{r  echo=FALSE,message=FALSE, warning=FALSE, gender_plot}

ggplot(gender,aes(x=as.factor(gender)))+
  geom_bar()+
    labs(x="Candidate gender",title = "Candidate Gender Bar plot")
  

table(gender$gender)
```

From this plot its clear the proportion of male candidates has been more than 
the female candidates in the presidential elections with 3 female candidates 
v/s 20 male candidates

```{r echo=FALSE,warning=FALSE,message=FALSE, candidate_party_plot_count}

candidate_party<-unique(data.table(ca_contrib),by=c("cand_nm"))
table(candidate_party$party)

qplot(x=party,data=candidate_party,xlab = "Political Party")+
  ggtitle("Candidate count per political party")

```

From this plot its clear that the number of candidates contesting the 
presidential election with the maximum being for Republican party with count
of 17 followed by Democratic having count of 5


I wanted to analyze the contributions made per political party

```{r echo=FALSE,warning=FALSE,message=FALSE, contrib_party_plot}

election_party_contrib<-summarise(group_by(ca_contrib,party),
                                  count =n())

election_party_contrib

ggplot(aes(x=party,y=count),data=election_party_contrib)+
  geom_bar(stat="identity")+
  xlab("Election Party")+
  ggtitle("Plotting contribution per party")+
  scale_y_log10()

```

Here the contribution had been maximum for Democratic candidates with count of 
1096563, followed by 2842 for Greens.


On analyzing the dataset, I found large number of negative 
contribution amounts for the contributors which on digging further I found 
were refunds when the candidate drops out of the race. So I summed the
contribution amount based on the contributor details and the candidate for
whom contribution has been made.There are still some negative values present
which might mean that some data is missing due to which the contribution
amounts have not been cancelled out in the cleaned data set.

```{r echo=FALSE,message=FALSE, warning=FALSE, plot_contribution_amounts}

#Summed the total contribution based on contibutor and candidate

ca_contrib_amt<-ca_contrib%>% 
  group_by(cand_nm,candidate_gender,contbr_nm,party,contbr_city,contbr_zip,
           contributor_gender,contbr_occupation,election_type) %>%
  summarise(total_contrib = sum(contb_receipt_amt))

contrib_density <-density(ca_contrib_amt$total_contrib)
plot(contrib_density,main="Contributor Density Amount")

summary(ca_contrib_amt$total_contrib)

```

The above contributor density plot still shows lot of negative contribution 
amount, which for the purpose of this investigation should be ignored. 

```{r echo=FALSE,message=FALSE, warning=FALSE, plot_contribution_amounts_limits}

qplot(x=total_contrib,data=ca_contrib_amt,binwidth=50,
      xlab = "Total contribution")+
  scale_x_continuous(limits = c(0,3000),breaks = seq(0,43000,1000))+
  ggtitle("Plot for contributions amount")

```

I transformed the total contribution visualization to get a better understanding
of the plot by re- arranging the axis from 0 to 3000 since maximum contribution 
appear in the range  of 0-3000$. The plot is positively skewed with an outlier 
at around 2800$. The maximum contribution appears in the range 0-500$.


```{r echo=FALSE,message=FALSE, warning=FALSE, plot_contb_occuption}

#Find the top 10 contributor occupation
contributor_occupation<-summarise(group_by(ca_contrib,contbr_occupation),
                                  count =n())
contributor_occupation<-arrange(contributor_occupation,desc(count))[1:10,]

contributor_occupation

ggplot(aes(x=contbr_occupation,y=count),data=contributor_occupation)+
  geom_bar(stat="identity")+
  xlab("Contributor occupation")+
  ggtitle("Plotting Frequent contributors occupation")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5,size = 10))

```

I wanted to determine the overall  occupation of contributors who contributed 
in the presidential elections. So the Bar plot above represents the top 10 
contributor's occupation  with the top most being Retired having count of 
260546.


```{r echo=FALSE,message=FALSE, warning=FALSE, plot_contb_employer}

#Find the top 10 contributor employer
contributor_employer<-summarise(group_by(ca_contrib,contbr_employer),
                                  count =n())
contributor_employer<-arrange(contributor_employer,desc(count))[1:10,]

contributor_employer

ggplot(aes(x=contbr_employer,y=count),data=contributor_employer)+
  geom_bar(stat="identity")+
  xlab("Contributor Employer")+
  ggtitle("Plotting frequent contributor's employers")+
  theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5,size = 10))


```

Also I want to determine who were the employers whose employees participated 
in the contribution. From this bar plot it shows that contributors 
having blank employement information have the highest count of contribution 
followed by the retired employees in this dataset. The reason for this
blank employee information I could think of could be because of either missing 
data/employee's didn't want to disclose their employer at the time of filing 
these contributions or probably they didn't have any employer!!


Similarly I want to determine the year with the maximum contributions made by
the contributors.

```{r  echo=FALSE,message=FALSE, warning=FALSE, contribution_per_year}

qplot(x=contrib_year,data=ca_contrib,xlab = "Contribution year")+
  ggtitle("Contributions based on year")

table(ca_contrib$contrib_year)

```

This shows that the maximum contribution from years 2013-2016 has been in year
2016. And though this data is for 2016 presidential election, there are some
insignificant count of contributions made in the year 2013-2014 which have
been identified while plotting this visualization.

```{r echo=FALSE,message=FALSE, warning=FALSE, boxplot_contribution_gender}

qplot(x = contributor_gender,y = total_contrib, 
      data = subset(ca_contrib_amt,!is.na(contributor_gender)),
      geom='boxplot',xlab = "Gender", ylab = "Contributions")+
  coord_cartesian(ylim = c(0,2000))+
  ggtitle("Contributions based on gender")

by(ca_contrib_amt$total_contrib,ca_contrib_amt$contributor_gender,summary)
```

This plot above shows the contribution made on the basis of gender by 
male/female contributors. I chose to ignore the negative contributions as well
as the contributors whose gender could not de determined from gender package. 
So I set the range of this boxplot to 0-2000 dollars.
This boxplot shows that for female contributors 25 percent contributed less 
than 78 dollars and for male contributors this amount is 80 dollars. 
The median for female and male contributors had been 200 dollars and 215 
dollars respectively.25% of female and male contributors contributed more than 
500 dollars and with the maximum being 21250 dollars for females and 27000
dollars for males.

# Univariate Analysis


### What is the structure of your dataset?

There were 304346 records of presidential election contribution in the year 
2016 from the state of California in this dataset with around 18 variables.
The factored variables in this data set are form type with 
values (SA17A,SA18,SB28A), Election type with values (Primary,General,Other,
Convention,Runoff,Special,Recount).

Other observations about the dataset :<br/>
* Most candidates are male<br/>
* Maximum number of contesting candidates were from Democratic party<br/>
* Maximum number of contributions had been for Democratic party candidates<br/>
* Most number of contribution amounts had been in the range 0-3000$<br/>
* "Retired" occupation had the maximum number of contributors<br/>
* Maximum contribution amount had been from male contributors.<br/>

### What is/are the main feature(s) of interest in your dataset?

The main feature that interests me in this dataset are the contribution amount
and the candidate for whom this contribution has been made.I'd like to examine
whatfeatures influences contributors to contribute to a political party. 
Probably the contributor has some affiliations to a political party or maybe 
some other variable combinations can be used to determine a predictive model.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

The candidate's political party, candidate name, election type, candidate's
gender, contributor's gender are some of the factors I could think of which
might influence a contributor to contribute for a particular candidate. On the
basis of my current analysis, I think political party plays a definitive role
in the contribution.

### Did you create any new variables from existing variables in the dataset?

Yes, I added a couple of variables in the dataset which would help me further
in my analysis. These variables were<br/>
* first_name and last_name: first and last name of candidate<br/>
* candidate_gender : gender of candidate<br/>
* party: political party of the candidate<br/>
* election_type: identified type of election from abbreviated election_type like
  Primary, General etc.<br/>
* contrib_year: Contribtion year extracted from contribution date<br/>
* contrib_month: contribution month extracted from contribution date<br/>


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

While analyzing thecontribution amounts, I found large number of negative 
contribution amounts which were basically refunds made to the contributors.
Since having negative contributions are not valid contributions.So to tidy this 
dataset, I summed the total contributions contribution amount based on 
contributor details and candidate contributed to in a variable called 
total_contrib in another dataset and plotted visualizations based on the same.
Also large number of columns had blank values associated with them, considering
which could lead to incorrect analysis, so I replaced these blank values with
"N/A"


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!