Financial Contributions to Presidential Campaigns by State by Ruchita Maheshwary
================================================================================

In this report, I analyzed the data set for around 1304500 contributions for 
presidential elections from the state of California.

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}

library(ggplot2)
library(gender)
library(tidyr)
library(dplyr)

```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <-read.csv('P00000001-CA.csv',row.names = NULL)
#matching correct column names with the data
colnames(df) <- c(colnames(df)[-1],"x")
df$x <- NULL

```

**Brief description about the dataset variables:**

* COMMITTEE ID- A 9-character alpha-numeric code assigned to a committee by the 
Federal Election Commission. 

* CANDIDATE ID - A 9-character alpha-numeric code assigned to a candidate by the
Federal Election Commission. 

* CANDIDATE NAME

* CONTRIBUTOR NAME - Reported name of the contributor.

* CONTRIBUTOR CITY - Reported city of the contributor.

* CONTRIBUTOR STATE - Reported state of the contributor.

* CONTRIBUTOR ZIP CODE - Reported zip code of the contributor.

* CONTRIBUTOR EMPLOYER - Reported employer of the contributor.

* CONTRIBUTOR OCCUPATION - Reported occupation of the contributor.

* CONTRIBUTION RECEIPT AMOUNT - Reported contribution amount.

* CONTRIBUTION RECEIPT DATE - Reported contribution receipt date.  
  The date format is DD-MMM-YYYY.

* RECEIPT DESCRIPTION - Additional information reported by the committee about 
  a specific contribution. 

* MEMO CODE - 'X' indicates the reporting committee has provided additional text
to describe a specific contribution.  See the MEMO TEXT.  

* MEMO TEXT - Additional information reported by the committee about a specific 
contribution.  

* FORM TYPE - Indicates what schedule and line number the reporting committee
reported a specific transaction.<br/>   

    * SA17A:  Form 3P Schedule A Line 17A<br/>
    * SA18:   Form 3P Schedule A Line 18<br/>
    * SB28A:  Form 3P Schedule B Line 28A<br/>  

* FILE NUMBER - A unique number assigned to a report and all its associated 
transactions.   

* TRANSACTION ID - A unique identifier permanently associated with each
itemization or transaction appearing in an FEC electronic file.

* ELECTION TYPE / PRIMARY-GENERAL INDICATOR - This code indicates the election 
for which the contribution was made. EYYYY (election plus election year)<br/>

    * P = Primary<br/>
    * G = General<br/>
    * O = Other<br/>
    * C = Convention<br/>
    * R = Runoff<br/>
    * S = Special<br/>
    * E = Recount<br/>


# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
names(df)

str(df)

summary(df)
```

```{r echo=FALSE, load_first_last_name}
#Load candidates by first and last name
df$cand_nm.new <- as.character(df$cand_nm)
#split candidate name by first and last name and identify unique candidate names
#in the data set
ca_contrib <- extract(df, cand_nm.new, c("last_name","first_name"), 
                      "(.*),\\s*(\\S*)\\s*.*", fill="left",perl=TRUE)
ca_candidates <- unique(ca_contrib$first_name)

```

```{r echo=FALSE, candidate_gender}
#Load gender of candidates
#retrieve gender of the candidates with year of birth between 1941-1971
gender <- gender(as.character(ca_candidates), c(1941, 1971),method = "ssa",
                 countries = "United States")
#join with the data set the gender data and remove additional columns
ca_contrib<-left_join(x=ca_contrib,y=gender,by=c("first_name"= "name"))
names(ca_contrib)[names(ca_contrib) == 'gender'] <- 'candidate_gender'
drop <- c("name","proportion_male","proportion_female","year_min","year_max")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
ca_contrib$candidate_gender <- as.factor(ca_contrib$candidate_gender)
```

```{r echo=FALSE, load_contributor_details}
#Repeat the same process for contributors
ca_contrib$contrib_nm.new <- as.character(ca_contrib$contbr_nm)
#retrieve contributor first and last name
ca_contrib <- extract(ca_contrib, contrib_nm.new, c("contrib_last_name",
                                                    "contrib_first_name"),
                      "(.*),\\s*(\\S*)\\s*.*", fill="left",perl=TRUE)
ca_contributors <- unique(ca_contrib$contrib_first_name)
#retrieve gender of the contributors with year of birth between 1922-1997
gender_contrib <- gender(as.character(ca_contributors), c(1922, 1997),
                         method = "ssa",countries = "United States")
#join with the data set the gender data and remove additional columns
ca_contrib<-left_join(x=ca_contrib,y=gender_contrib,
                      by=c("contrib_first_name"= "name"))
names(ca_contrib)[names(ca_contrib) == 'gender'] <- 'contributor_gender'
drop <- c("contrib_last_name","contrib_first_name","name",
          "proportion_male","proportion_female","year_min","year_max")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
ca_contrib$contributor_gender <- as.factor(ca_contrib$contributor_gender)

```

```{r echo=FALSE, remove_columns}
#Removing non-usable columns from data frame like committe id,receipt 
#description and redundant columns like contributor state etc
drop <- c("cmte_id","contbr_st","receipt_desc","memo_cd",
          "memo_text","tran_id")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]

```

```{r echo=FALSE, load_candidate_party}
#Add column of political party for each presidential election candidate based on
#candidate last name

candidateParty <- function(x) {
    if(is.na(x)){
      NA
    }else if(x == "Clinton" | x == "Sanders" | x == "O'Malley" | x=="Lessig" 
             | x=="Webb"){
      "Democratic"
    }else if(x == "Johnson"){
      "Libertarian"
    }else if(x=="Stein"){
      "Green"
    }else if(x=="McMullin") {
      "Independents"
    }else {
      "Republican"
    }
}

ca_contrib$party <- apply(ca_contrib['last_name'],1,candidateParty)

```

```{r echo=FALSE, load_election_type}

#Add election type description for "election_tp" column

electionType <- function(x) {
  if(is.na(x))  {
    NA
  }else if(substr(x, 1, 1)=="P")  {
    "Primary"
  }else if(substr(x, 1, 1)=="G")  {
    "General"
  }else if(substr(x, 1, 1)=="O")  {
    "Other"
  }else if(substr(x, 1, 1)=="C")  {
    "Convention"
  }else if(substr(x, 1, 1)=="R")  {
    "Runoff"
  }else if(substr(x, 1, 1)=="S")  {
    "Special"
  }else {
    "Recount"
  }
}

ca_contrib$election_type <- apply(ca_contrib['election_tp'],1,electionType)
drop <- c("election_tp")
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% drop)]
  
```


```{r echo=FALSE, retrieve_contribution_dates}

#Obtain contribution month/year
ca_contrib$contribution_date <- as.Date(ca_contrib$contb_receipt_dt,"%d-%B-%y")
ca_contrib$contrib_year = as.numeric(format(ca_contrib$contribution_date,format="%Y"))
ca_contrib$contrib_month = months(as.Date(ca_contrib$contribution_date))
ca_contrib = ca_contrib[,!(names(ca_contrib) %in% c("contribution_date"))]

```


```{r  echo=FALSE, contribution_amounts}
#On analyzing the data I found some negative contributions as well which were 
#basically refunds (typically following a candidate dropping out of the race). 
#So will subset the data into positive and negative contributions

positive_ca_contrib_amt <-subset(ca_contrib,contb_receipt_amt>0)
negative_ca_contrib_amt <-subset(ca_contrib,contb_receipt_amt<=0)

```

After cleaning the data our dataset contains 1304346 observations of 19 variables.

Plot for male/female candidate count in the presidential election
```{r  echo=FALSE, gender_plot}
qplot(gender,data=gender)

```

Plot for contribution amount 
```{r echo=FALSE,message=FALSE, warning=FALSE, plot_contribution_amounts}

#summary(positive_ca_contrib_amt$contb_receipt_amt)

qplot(x=contb_receipt_amt,data=positive_ca_contrib_amt,binwidth=100)+
 scale_x_continuous(limits=c(0,11000),breaks=seq(0,110000,2000))


```

```{r echo=FALSE,message=FALSE, warning=FALSE, plot_log_contribution_amounts}

ggplot(aes(x=contb_receipt_amt),data=positive_ca_contrib_amt)+
  geom_histogram(binwidth=100)+
  scale_y_sqrt()

```



```{r}
#by(positive_ca_contrib_amt$contb_receipt_amt,positive_ca_contrib_amt$candidate_gender,summary)
qplot(x=contb_receipt_amt,data=positive_ca_contrib_amt,binwidth=200)+

  facet_wrap(~candidate_gender)
```

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!